---
title: "Hands-on Exercise 2"
editor: visual
---

## 1 Overview

This hands-on exercise is on **global and local measures of spatial autocorrelation**.

### 1.1 The analytical question

In this case study, we are interested to examine the spatial pattern of a selected development indicator (i.e. GDP per capita) of Hunan Provice, People Republic of China.

## 2 Getting Ready

### 2.1 Load the required packages:

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse)
```

### 2.2 Import shapefile

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```

### 2.3 Import csv file

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

### 2.4 Perform relational join and visualisation

```{r}
hunan <- left_join(hunan,hunan2012)
```

Prepare a basemap and choropleth map showing the distribution of GDPPC 2012:

```{r}
equal <- tm_shape(hunan) +
  tm_fill("GDPPC",
          n = 5,
          style = "equal") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_fill("GDPPC",
          n = 5,
          style = "quantile") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Equal quantile classification")

tmap_arrange(equal, 
             quantile, 
             asp=1, 
             ncol=2)
```

## 3 Global Spatial Autocorrelation

### 3.1 Compute Contiguity Spatial Weights

```{r}
wm_q <- poly2nb(hunan, 
                queen=TRUE)
summary(wm_q)
```

The summary report above shows that there are 88 area units in Hunan. The most connected area unit has 11 neighbours. There are two area units with only one heighbours.

### 3.2 Row-standardised weights matrix

```{r}
rswm_q <- nb2listw(wm_q, 
                   style="W", 
                   zero.policy = TRUE)
rswm_q
```

The input of *nb2listw()* must be an object of class **nb**. The syntax of the function has two major arguments, namely style and zero.poly.

-   *style* can take values "W", "B", "C", "U", "minmax" and "S". B is the basic binary coding, W is row standardised (sums over all links to n), C is globally standardised (sums over all links to n), U is equal to C divided by the number of neighbours (sums over all links to unity), while S is the variance-stabilizing coding scheme proposed by Tiefelsdorf et al. 1999, p. 167-168 (sums over all links to n).

-   If *zero policy* is set to TRUE, weights vectors of zero length are inserted for regions without neighbour in the neighbours list. These will in turn generate lag values of zero, equivalent to the sum of products of the zero row t(rep(0, length=length(neighbours))) %\*% x, for arbitrary numerical vector x of length length(neighbours). The spatially lagged value of x for the zero-neighbour region will then be zero, which may (or may not) be a sensible choice.

### 3.3 Global Spatial Autocorrelation: Moran's I

We will use ***moran.test()*** of **spdep** to perform Moran's I statistics testing.

```{r}
moran.test(hunan$GDPPC, 
           listw=rswm_q, 
           zero.policy = TRUE, 
           na.action=na.omit)
```

Since p\<0.05, we reject null hypothesis at 5% significance level. Since Moran I statistic is positive (\>0), this means that observations tend to be similar i.e. clustered.

#### Computing Monte Carlo Moran's I

We use ***moran.mc()*** of **spdep** to perform permutation test for Moran's I statistic (1000 simulations).

```{r}
set.seed(1234)
bperm= moran.mc(hunan$GDPPC, 
                listw=rswm_q, 
                nsim=999, 
                zero.policy = TRUE, 
                na.action=na.omit)
bperm
```

Since p\<0.05, we reject null hypothesis at 5% significance level. Since Moran I statistic is positive (\>0), this means that observations tend to be similar i.e. clustered.

#### Visualising Monte Carlo Moran's I

```{r}
mean(bperm$res[1:999])
var(bperm$res[1:999])
summary(bperm$res[1:999])
```

```{r}
hist(bperm$res, 
     freq=TRUE, 
     breaks=20, 
     xlab="Simulated Moran's I")
abline(v=0, 
       col="red") 
```

### 3.4 Global Spatial Autocorrelation: Geary's C

We will use ***geary.test()*** of **spdep** to perform Geary's C test for spatial autocorrelation.

```{r}
geary.test(hunan$GDPPC, listw=rswm_q)
```

![](data:image/png;base64,iVBORw0KGgoAAAANSUhEUgAAACgAAAAaCAYAAADFTB7LAAAAa0lEQVR42u3OywnAIBBAwcXSUoCW5D11xDoNCBGNv0MOecJOBSOi1OZMsJ4dvFxEJ1OQnMxBarIKEpNNkJbsBknJYZCSnAYJyVVQziNig7/nZkFEbhTE5HpBVO4dxOXKIDL3BLG5BJ1T6rsbMfep2CaMN00AAAAASUVORK5CYII= "Run Current Chunk")

Since p\<0.05, we reject null hypothesis at 5% significance level. Since Geary C statistic is small (\<1), this means that observations tend to be similar i.e. clustered.

#### Computing Monte Carlo Geary's C

We use ***geary.mc()*** of **spdep** to perform permutation test for Geary's C statistic (1000 simulations).

```{r}
set.seed(1234)
bperm=geary.mc(hunan$GDPPC, 
               listw=rswm_q, 
               nsim=999)
bperm
```

Since p\<0.05, we reject null hypothesis at 5% significance level. Since Geary C statistic is small (\<1), this means that observations tend to be similar i.e. clustered.

#### Visualising Monte Carlo Geary's C

```{r}
mean(bperm$res[1:999])
var(bperm$res[1:999])
summary(bperm$res[1:999])
```

```{r}
hist(bperm$res, freq=TRUE, breaks=20, xlab="Simulated Geary c")
abline(v=1, col="red") 
```

## 4 Spatial Correlogram

### 4.1 Compute Moran's I correlogram

```{r}
MI_corr <- sp.correlogram(wm_q, 
                          hunan$GDPPC, 
                          order=6, 
                          method="I", 
                          style="W")
plot(MI_corr)
```

By plotting the output might not allow us to provide complete interpretation. This is because not all autocorrelation values are statistically significant. Hence, it is important for us to examine the full analysis report by printing out the analysis results as in the code chunk below.

```{r}
print(MI_corr)
```

### 4.2 Compute Geary's C correlogram and plot

```{r}
GC_corr <- sp.correlogram(wm_q, 
                          hunan$GDPPC, 
                          order=6, 
                          method="C", 
                          style="W")
plot(GC_corr)
```

```{r}
print(GC_corr) 
```

## 5 Cluster and Outlier Analysis

### 5.1 Computing local Moran's I

We will use ***localmoran()*** of **spdep** to compute local Moran's I.

```{r}
fips <- order(hunan$County)
localMI <- localmoran(hunan$GDPPC, rswm_q)
head(localMI)
```

*localmoran()* function returns a matrix of values whose columns are:

-   Ii: the local Moran's I statistics

-   E.Ii: the expectation of local moran statistic under the randomisation hypothesis

-   Var.Ii: the variance of local moran statistic under the randomisation hypothesis

-   Z.Ii:the standard deviate of local moran statistic

-   Pr(): the p-value of local moran statistic

```{r}
printCoefmat(data.frame(localMI[fips,], row.names=hunan$County[fips]), check.names=FALSE)
```

#### Mapping local Moran's I Values

```{r}
hunan.localMI <- cbind(hunan,localMI) %>%
  rename(Pr.Ii = Pr.z....E.Ii..)
```

```{r}
tm_shape(hunan.localMI) +
  tm_fill(col = "Ii", 
          style = "pretty",
          palette = "RdBu",
          title = "local moran statistics") +
  tm_borders(alpha = 0.5)
```

#### Mapping local Moran's I p-values

```{r}
tm_shape(hunan.localMI) +
  tm_fill(col = "Pr.Ii", 
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette="-Blues", 
          title = "local Moran's I p-values") +
  tm_borders(alpha = 0.5)
```

#### Mapping both local Moran's I values and p-values

```{r}
localMI.map <- tm_shape(hunan.localMI) +
  tm_fill(col = "Ii", 
          style = "pretty", 
          title = "local moran statistics") +
  tm_borders(alpha = 0.5)

pvalue.map <- tm_shape(hunan.localMI) +
  tm_fill(col = "Pr.Ii", 
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette="-Blues", 
          title = "local Moran's I p-values") +
  tm_borders(alpha = 0.5)

tmap_arrange(localMI.map, pvalue.map, asp=1, ncol=2)
```

***From the graph on the left, the orange region represents outlier while the green regions are clusters.***

***However, for the outlier region, as there are only 3 neighbours, it is not statistically significant as shown by the p-value on the right.***

## 6 Creating a LISA Cluster Map

The LISA Cluster Map shows the significant locations color coded by type of spatial autocorrelation. The first step before we can generate the LISA cluster map is to plot the Moran scatterplot.

### 6.1 Plot Moran scatterplot

The Moran scatterplot is an illustration of the relationship between the values of the chosen attribute at each location and the average value of the same attribute at neighboring locations.

```{r}
nci <- moran.plot(hunan$GDPPC, rswm_q,
                  labels=as.character(hunan$County), 
                  xlab="GDPPC 2012", 
                  ylab="Spatially Lag GDPPC 2012")
```

The plot is split in 4 quadrants. The top right corner belongs to areas that have high GDPPC and are surrounded by other areas that have the average level of GDPPC. This are the high-high locations in the lesson slide.

### 6.2 Plot Moran scatterplot with standardised variable

We use ***scale()*** to center and scale the variable (i.e. standardisation - subtract mean and divide by standard deviation).

The **as.vector()** ensures that the resulting data type is a vector, that map neatly into our dataframe.

```{r}
hunan$Z.GDPPC <- scale(hunan$GDPPC) %>% as.vector 
```

```{r}
nci2 <- moran.plot(hunan$Z.GDPPC, rswm_q,
                   labels=as.character(hunan$County),
                   xlab="z-GDPPC 2012", 
                   ylab="Spatially Lag z-GDPPC 2012")
```

### 6.3 Preparing LISA map classes

```{r}
quadrant <- vector(mode="numeric",length=nrow(localMI))
hunan$lag_GDPPC <- lag.listw(rswm_q, hunan$GDPPC)
DV <- hunan$lag_GDPPC - mean(hunan$lag_GDPPC)     
LM_I <- localMI[,1] - mean(localMI[,1])     
signif <- 0.05    
```

Define the high-high, low-low, low-high, and high-low categories:

```{r}
quadrant[DV <0 & LM_I>0] <- 1
quadrant[DV >0 & LM_I<0] <- 2
quadrant[DV <0 & LM_I<0] <- 3  
quadrant[DV >0 & LM_I>0] <- 4    
```

Place non-significant Moran in category 0.

```{r}
quadrant[localMI[,5]>signif] <- 0
```

### 6.4 Plotting LISA map

```{r}
hunan.localMI$quadrant <- quadrant
colors <- c("#ffffff", "#2c7bb6", "#abd9e9", "#fdae61", "#d7191c")
clusters <- c("insignificant", "low-low", "low-high", "high-low", "high-high")

tm_shape(hunan.localMI) +
  tm_fill(col = "quadrant", 
          style = "cat", 
          palette = colors[c(sort(unique(quadrant)))+1], 
          labels = clusters[c(sort(unique(quadrant)))+1],
          popup.vars = c("")) +
  tm_view(set.zoom.limits = c(11,17)) +
  tm_borders(alpha=0.5)
```

```{r}
gdppc <- qtm(hunan, "GDPPC")

hunan.localMI$quadrant <- quadrant
colors <- c("#ffffff", "#2c7bb6", "#abd9e9", "#fdae61", "#d7191c")
clusters <- c("insignificant", "low-low", "low-high", "high-low", "high-high")

LISAmap <- tm_shape(hunan.localMI) +
  tm_fill(col = "quadrant", 
          style = "cat", 
          palette = colors[c(sort(unique(quadrant)))+1], 
          labels = clusters[c(sort(unique(quadrant)))+1],
          popup.vars = c("")) +
  tm_view(set.zoom.limits = c(11,17)) +
  tm_borders(alpha=0.5)

tmap_arrange(gdppc, LISAmap, asp=1, ncol=2)
```

## 7 Hot Spot and Cold Spot Area Analysis

### 7.1 Getis and Ord's G-Statistics

The Getis and Ord's G-Statistics looks at neighbours within a defined proximity to identify where either high or low values clutser spatially.

Here, statistically significant hot-spots are recognised as areas of high values where other areas within a neighbourhood range also share high values too.

The analysis consists of three steps:

-   Deriving spatial weight matrix

-   Computing Gi statistics

-   Mapping Gi statistics

### 

7.2 Deriving distance-based weight matrix

#### Deriving the centroid

```{r}
longitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[1]])
latitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[2]])
coords <- cbind(longitude, latitude)
```

#### Determining the cut-off distance

```{r}
#coords <- coordinates(hunan)
k1 <- knn2nb(knearneigh(coords))
k1dists <- unlist(nbdists(k1, coords, longlat = TRUE))
summary(k1dists)
```

The summary report shows that the largest first nearest neighbour distance is 61.79 km, so using this as the upper threshold gives certainty that all units will have at least one neighbour.

#### Computing fixed distance weight matrix

```{r}
wm_d62 <- dnearneigh(coords, 0, 62, longlat = TRUE)
wm_d62
```

```{r}
wm62_lw <- nb2listw(wm_d62, style = 'B')
summary(wm62_lw)
```

### 7.3 Computing adaptive distance weight matrix

```{r}
knn <- knn2nb(knearneigh(coords, k=8))
knn
```

```{r}
knn_lw <- nb2listw(knn, style = 'B')
summary(knn_lw)
```

### 7.4 Computing Gi statistics

#### Gi statistics using fixed distance

```{r}
fips <- order(hunan$County)
gi.fixed <- localG(hunan$GDPPC, wm62_lw)
gi.fixed
```

The output of localG() is a vector of G or Gstar values, with attributes "gstari" set to TRUE or FALSE, "call" set to the function call, and class "localG".

The Gi statistics is represented as a Z-score. Greater values represent a greater intensity of clustering and the direction (positive or negative) indicates high or low clusters.

```{r}
hunan.gi <- cbind(hunan, as.matrix(gi.fixed)) %>%
  rename(gstat_fixed = as.matrix.gi.fixed.)
```

The code chunk above performs three tasks. First, it convert the output vector (i.e. *gi.fixed*) into r matrix object by using *as.matrix()*. Next, *cbind()* is used to join hunan\@data and *gi.fixed* matrix to produce a new SpatialPolygonDataFrame called *hunan.gi*. Lastly, the field name of the gi values is renamed to *gstat_fixed* by using *rename()*.

#### Mapping Gi values with fixed distance weights

```{r}
gdppc <- qtm(hunan, "GDPPC")

Gimap <-tm_shape(hunan.gi) +
  tm_fill(col = "gstat_fixed", 
          style = "pretty",
          palette="-RdBu",
          title = "local Gi") +
  tm_borders(alpha = 0.5)

tmap_arrange(gdppc, Gimap, asp=1, ncol=2)
```

#### Gi statistics using adaptive distance

```{r}
fips <- order(hunan$County)
gi.adaptive <- localG(hunan$GDPPC, knn_lw)
hunan.gi <- cbind(hunan, as.matrix(gi.adaptive)) %>%
  rename(gstat_adaptive = as.matrix.gi.adaptive.)
```

#### Mapping Gi values with adaptive distance weights

```{r}
gdppc<- qtm(hunan, "GDPPC")

Gimap <- tm_shape(hunan.gi) + 
  tm_fill(col = "gstat_adaptive", 
          style = "pretty", 
          palette="-RdBu", 
          title = "local Gi") + 
  tm_borders(alpha = 0.5)

tmap_arrange(gdppc, 
             Gimap, 
             asp=1, 
             ncol=2)
```
